hbm_paper_results

### 2024-04-11 paper_lbcc_results.Rmd
parsing through

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel, ggrain, skimr, gam, ggeffects)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

Combat-GAM Centiles - site-level batch correction

@@

@@

tbl.df <- as.data.frame(table(cent_deltas_df$site, cent_deltas_df$study))
```

@@

```{r}
delta_n.sex <- cent_deltas_df_long %>%
  mutate(abs.delta = abs(delta)) %>%
  group_by(sex, study) %>%
  summarise(sex_n = n(),
            mean_abs.delta=mean(abs.delta),
            mean_delta=mean(delta),
            mean_age_yrs=mean(age_days)/365.25) #get number males and females in each site

@@

look at max abs. delta across sites
```{r}
delta_sum.df_merged %>%
  ggplot() +
  geom_histogram(aes(x=study_max_abs.delta*100))

delta_sum.df_merged %>%
  ggplot(aes(y=study_max_abs.delta*100, x=study_n)) +
  geom_point() +
  geom_smooth(method="lm")

delta_sum.df_merged %>%
  ggplot(aes(y=study_max_abs.delta*100, x=study_mean_age_yrs)) +
  geom_point() +
  geom_smooth(method="lm")

delta_sum.df_merged %>%
  ggplot(aes(y=study_max_abs.delta*100, x=sex_prop_Female)) +
  geom_point() +
  geom_smooth(method="lm")
```
#Batch Effects

@@

```{r}
gam.feat_to_test <- intersect(pheno_list, names(gam_centile_df.clean))

@@

df_gam_feat_aov.study <- df_gam_feat_aov %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(gam.feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gam_feat_aov.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

@@

```{r}
#clean predictions
gamlss_centile_df.clean <- gamlss_centile_df %>%
  mutate(study=sub("\\_.*", "", site),
         age_years=age_days/365.25) %>%
  dplyr::select(!contains(".z") & !ends_with("_total") & !contains("TBV")) #drop z-scores and non-centiles
```

@@

df_gamlss_feat_aov.study <- df_gamlss_feat_aov %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gamlss_feat_aov.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

@@

@@

@@

gamlss_centile_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_cent, color=site), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_cent), method="lm") +
  theme(legend.position = "none") +
  facet_wrap(~study)

gamlss_centile_df.mean %>%
  ggplot() +
  geom_boxplot(aes(x=site, y=mean_cent, fill=study)) + 
  theme(axis.text.x = element_text(angle=90), legend.position = "none")

@@

#Centile Fans
Look at brain charts for global vols

```{r}
gamlss.GMV.mod <- readRDS(paste0(der_path, "/GMV_lifespan-CN-imp-sites-euler-log-cf-cf.gamlss-batch.study-data_mod.rds"))

```

@@

NS

@@

```{r}
# age
summary(lm(study_mean_abs.delta ~ study_mean_age_yrs, data=delta_sum.df_merged))
summary(lm(study_mean_delta ~ study_mean_age_yrs, data=delta_sum.df_merged))

#forward selection for site_mean_abs.delta
summary(lm(study_mean_abs.delta ~ study_mean_age_yrs + study_n, data=delta_sum.df_merged)) #small effect but remains significant/grows stronger
summary(lm(study_mean_abs.delta ~ study_mean_age_yrs + study_n + sex_prop_Female, data=delta_sum.df_merged)) #small effect but remains significant/grows stronger
full.delta.age.fit <- lm(study_mean_abs.delta*100 ~ study_mean_age_yrs + study_n + sex_prop_Female, data=delta_sum.df_merged) #scaling to centile space
summary(full.delta.age.fit)

#forward selection for site_mean_delta
summary(lm(study_mean_delta ~ study_mean_age_yrs + study_n, data=delta_sum.df_merged)) #small effect but remains significant/grows stronger
summary(lm(study_mean_delta ~ study_mean_age_yrs + study_n + sex_prop_Female, data=delta_sum.df_merged)) #small effect but remains significant/grows stronger
full.age.fit <- lm(study_mean_delta*100 ~ study_mean_age_yrs + study_n + sex_prop_Female, data=delta_sum.df_merged) #scaling to centile space
summary(full.age.fit)
```

test w/o babies < 2
```{r}
delta_sum.df_merged_noinf <- delta_sum.df_merged %>%
  dplyr::filter(study_mean_age_yrs >= 2)

nrow(delta_sum.df_merged) - nrow(delta_sum.df_merged_noinf)

summary(lm(study_mean_abs.delta ~ study_mean_age_yrs, data=delta_sum.df_merged_noinf))
summary(lm(study_mean_delta ~ study_mean_age_yrs, data=delta_sum.df_merged_noinf))

all_means.df.long %>%
  dplyr::filter(study_mean_age_yrs >= 2) %>%
  ggplot(aes(x=study_mean_age_yrs, y=mean_cent, color=combat, size=study_n)) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
delta_sum.df_merged %>%
  ggplot(aes(x=site_mean_age_yrs, y=site_mean_delta)) +
  geom_count(aes(size=site_n)) +
  geom_smooth(method="lm")
delta_sum.df_merged %>%
  ggplot(aes(x=site_mean_age_yrs, y=site_mean_abs.delta)) +
  geom_count(aes(size=site_n)) +
  geom_smooth(method="lm")

@@

are studies w older subjects changing more because there are fewer of them?
check for association between study size and age
```{r}
summary(lm(site_mean_age_yrs ~ site_n, data=delta_sum.df_merged)) #study size is associated with age, but assoc. remains when controlling for size
```

@@

delta_sum.df_merged %>%
  ggplot() +
  geom_histogram(aes(x=site_mean_age_yrs)) +
  labs(title="site distribution across age")
```

@@

```{r}
gam_centile_site.mean.df <- gam_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gam_cent=mean(centile))

gam.prop.f.test <- lm(mean_gam_cent ~ sex_prop_Female + site_n + site_mean_age, data=gam_centile_site.mean.df2)
summary(gam.prop.f.test)
sjPlot::plot_model(gam.prop.f.test, type = "pred", terms = "sex_prop_Female")

ggplot(gam_centile_site.mean.df2, aes(x=sex_prop_Female, y=mean_gam_cent)) +
  geom_point() +
  geom_smooth(method="lm")
```

@@

```{r}
gamlss_centile_site.mean.df <- gamlss_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(site) %>%
  summarize(mean_gamlss_cent=mean(centile))

gamlss.prop.f.test <- lm(mean_gamlss_cent ~ sex_prop_Female + site_n + site_mean_age, data=gamlss_centile_site.mean.df2)
summary(gamlss.prop.f.test)
sjPlot::plot_model(gamlss.prop.f.test, type = "pred", terms = "sex_prop_Female")

ggplot(gamlss_centile_site.mean.df2, aes(x=sex_prop_Female, y=mean_gamlss_cent)) +
  geom_point() +
  geom_smooth(method="lm")
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(data.table, ggplot2, tidyverse, ggseg, ggsegExtra, paletteer, ggpubr, gridExtra, lemon, parallel, rstatix, lme4, interactions, ggrepel, ggrain, skimr, gam, ggeffects, effectsize)
#Sys.setenv("MC_CORES"=10L)
options("mc.cores")

@@

Combat-GAM Centiles - study-level batch correction w/o infants

@@

```{r}
min(gam_centile_df$age_years) == min(gamlss_centile_df$age_years)
min(gamlss_centile_df$age_years)
```

@@

tbl.df <- as.data.frame(table(cent_deltas_df$site, cent_deltas_df$study))
min(gam_centile_df$age_years) == min(cent_deltas_df$age_years)
```

@@

```{r}
delta_n.sex <- cent_deltas_df_long %>%
  mutate(abs.delta = abs(delta)) %>%
  group_by(sex, study) %>%
  summarise(sex_n = n(),
            mean_abs.delta=mean(abs.delta),
            mean_delta=mean(delta),
            mean_age_yrs=mean(age_days)/365.25) #get number males and females in each study

@@

#Batch Effects

@@

```{r}
gam.feat_to_test <- intersect(pheno_list, names(gam_centile_df.clean))
length(gam.feat_to_test)

@@

df_gam_feat_aov.study <- df_gam_feat_aov %>%
  group_by(pheno) %>%
  mutate(etasq = sumsq / sum(sumsq)) %>%
  ungroup() %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(gam.feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gam_feat_aov.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

df_gam_feat_aov.study %>%
  ggplot() +
  geom_histogram(aes(x=etasq, fill=sig_fdr), alpha=0.5)

@@

```{r}
#clean predictions
gamlss_centile_df.clean <- gamlss_centile_df %>%
  mutate(study=sub("\\_.*", "", site),
         age_years=age_days/365.25) %>%
  dplyr::select(!contains(".z")) #drop z-scores and non-centiles & !ends_with("_total") & !contains("TBV")
```

```{r}
ggplot(gamlss_centile_df.clean) +
  geom_point(aes(x=age_days, y=rh_Vol_insula, color=study), alpha=0.5) +
  theme(legend.position="none")
```

@@

df_gamlss_feat_aov.study <- df_gamlss_feat_aov %>%
  group_by(pheno) %>%
  mutate(etasq = sumsq / sum(sumsq)) %>%
  ungroup() %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::mutate(p.val_fdr = p.adjust(p.value, method="fdr", n = length(feat_to_test)), #apply fdr correction across all phenos tested
                  sig_fdr = case_when(p.val_fdr < 0.05 ~ TRUE,
                                      p.val_fdr >= 0.05 ~ FALSE))
df_gamlss_feat_aov.study %>%
  ggplot() +
  geom_bar(aes(y=sig_fdr))

df_gamlss_feat_aov.study %>%
  ggplot() +
  geom_histogram(aes(x=etasq, fill=sig_fdr))

@@

```{r}
df_gamlss_feat_aov.study$combat <- "gamlss"
df_gam_feat_aov.study$combat <- "gam"

feat_aov.df <- full_join(df_gamlss_feat_aov.study, df_gam_feat_aov.study)

feat_aov.df <- feat_aov.df %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

feat_aov.df %>%
  ggplot() +
  geom_point(aes(x=etasq, y=pheno, color=combat, shape=sig_fdr), alpha=0.6) +
  facet_wrap(~pheno_cat, scales="free")
```

@@

```{r}
ggplot(gamlss_centile_df.mean) +
  geom_point(aes(x=(age_days/365.25), y=mean_cent, color=study), alpha=0.5) +
  theme(legend.position="none")

ggplot(gamlss_centile_df.clean) +
  geom_point(aes(x=age_days, y=lh_Vol_inferiortemporal, color=study), alpha=0.2) +
  theme(legend.position="none")
```

@@

gamlss_centile_df.mean %>%
  ggplot() +
  geom_point(aes(x=age_days/365.25, y=mean_cent, color=study), alpha=0.5) +
  geom_smooth(aes(x=age_days/365.25, y=mean_cent, color=study), method="lm") +
  theme(legend.position = "none")

@@

@@

@@

```{r}
# age
summary(lm(study_mean_abs.delta ~ study_mean_age_yrs, data=delta_sum.df_merged))
summary(lm(study_mean_delta ~ study_mean_age_yrs, data=delta_sum.df_merged))
```

```{r}
delta_sum.df_merged %>%
  ggplot(aes(x=study_mean_age_yrs, y=study_mean_delta)) +
  geom_count(aes(size=study_n)) +
  geom_smooth(method="lm")
delta_sum.df_merged %>%
  ggplot(aes(x=study_mean_age_yrs, y=study_mean_abs.delta)) +
  geom_count(aes(size=study_n)) +
  geom_smooth(method="lm")

@@

are studies w older subjects changing more because there are fewer of them?
check for association between study size and age
```{r}
summary(lm(study_mean_age_yrs ~ study_n, data=delta_sum.df_merged)) #study size is associated with age, but assoc. remains when controlling for size
```

@@

delta_sum.df_merged %>%
  ggplot() +
  geom_histogram(aes(x=study_mean_age_yrs)) +
  labs(title="study distribution across age")
```

@@

```{r}
gam_centile_study.mean.df <- gam_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(study) %>%
  summarize(mean_gam_cent=mean(centile),
            mean_gam_abs_cent=mean(abs(centile)))

gam_centile_study.mean.df2 <- left_join(gam_centile_study.mean.df, delta_sum.df_merged)

gam.prop.f.test <- lm(mean_gam_cent ~ sex_prop_Female + study_n + study_mean_age, data=gam_centile_study.mean.df2)
summary(gam.prop.f.test)
gam.prop.f.test2 <- lm(mean_gam_abs_cent ~ sex_prop_Female + study_n + study_mean_age, data=gam_centile_study.mean.df2)
summary(gam.prop.f.test2)

ggplot(gam_centile_study.mean.df2, aes(x=sex_prop_Female, y=mean_gam_cent)) +
  geom_point() +
  geom_smooth(method="lm")
```

@@

```{r}
gamlss_centile_study.mean.df <- gamlss_centile_df.clean %>%
  pivot_longer(cols=any_of(pheno_list), names_to="pheno", values_to = "centile") %>%
  group_by(study) %>%
  summarize(mean_gamlss_cent=mean(centile),
            mean_gamlss_abs_cent=mean(abs(centile)))

gamlss_centile_study.mean.df2 <- left_join(gamlss_centile_study.mean.df, delta_sum.df_merged)

gamlss.prop.f.test <- lm(mean_gamlss_cent ~ sex_prop_Female + study_n + study_mean_age, data=gamlss_centile_study.mean.df2)
summary(gamlss.prop.f.test)
gamlss.prop.f.test2 <- lm(mean_gamlss_abs_cent ~ sex_prop_Female + study_n + study_mean_age, data=gamlss_centile_study.mean.df2)
summary(gamlss.prop.f.test2)

ggplot(gamlss_centile_study.mean.df2, aes(x=sex_prop_Female, y=mean_gamlss_cent)) +
  geom_point() +
  geom_smooth(method="lm")
```
